---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Data is available online, as given in the assignment.


Code such as this, would be preferable:

>
> data <- fread("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip")
>

However, there are some unsolved memory-related issues with RStudio version 0.99.441, latest version for R version 3.2.0 (2015-04-16) on the x86_64-apple-darwin13.4.0 platform, which throws errors when attempting to unzip the data on the fly with read.csv, comparable to [documented issues](http://stackoverflow.com/questions/22643372/embedded-nul-in-string-error-when-importing-huge-csv-with-fread).


Parsing and visuals require the packages dplyr, lubridate and lattice.



```{r, echo=TRUE}
require(dplyr)
require(lubridate)
require(lattice)

data <- read.csv(file = "activity.csv", as.is = TRUE)
data$date <- ymd(data$date)

```




## What is mean total number of steps taken per day?

```{r, echo=TRUE}
perdiem <- data %>%
  group_by(date) %>%
  summarise(allsteps = sum(steps)) %>%
  filter(allsteps > 0)

hist(perdiem$allsteps)

```

The median number of steps taken per day is 
```{r, echo=TRUE}
median(perdiem$allsteps)
```
and the mean is 
```{r, echo=TRUE}
mean(perdiem$allsteps)
```

In these calculations, missing values have been ignored.


## What is the average daily activity pattern?

```{r, echo=TRUE}
perinterval <- data %>%
  group_by(interval) %>%
  summarise(steps = mean(steps, na.rm = TRUE)) 

plot(perinterval, type = "l")

```

The 5 minute interval with most steps, is 
```{r, echo=TRUE}
perinterval$interval[max(perinterval$steps)]
```


## Imputing missing values

There are a number of days/intervals where there are missing values for the steps variable, coded as NA and ignored up till now.

There are exactly 
```{r, echo=TRUE}
sum(is.na(data$steps))
```
missing measurements.

We can fill the gaps with the mean for that 5-minute interval, as calculated earlier.

```{r, echo=TRUE}
newdata <- data %>%
  filter(is.na(steps) == TRUE) %>%
  select(date, interval)

newdata <- inner_join(x = newdata, y = perinterval, by = "interval")

data <- filter(data, is.na(steps) == FALSE)
data <- rbind(data, newdata)

perdiem <- data %>%
  group_by(date) %>%
  summarise(allsteps = sum(steps)) %>%
  filter(allsteps > 0)

hist(perdiem$allsteps)

```

The median number of steps taken per day is now identical to the mean, as an effect of manipulating the dataset:
```{r, echo=TRUE}
median(perdiem$allsteps)
mean(perdiem$allsteps)
```




## Are there differences in activity patterns between weekdays and weekends?

One can expect differences between weekdays and weekends. This can be shown visually.

```{r, echo=TRUE}

moredata <- data %>%
  mutate(weekday = wday(date))
moredata$weekday[moredata$weekday == 1] <- "weekend"
moredata$weekday[moredata$weekday == 7] <- "weekend"
moredata$weekday[nchar(moredata$weekday) == 1] <- "weekday"
moredata$weekday <- as.factor(moredata$weekday)

groupedperinterval <- moredata %>%
  group_by(weekday, interval) %>%
  summarise(steps = mean(steps, na.rm = TRUE)) 

xyplot(steps ~ interval | weekday, data = groupedperinterval, type = "l")

```


